---
title: "re_tidy"
author: "David Fäh 23-106-396"
output: html_document
date: "2025-03-01"
---

```{r setup, warning=FALSE, message=FALSE}

# load libraries
library(tidyverse)

# load data
co2_impact_data <- read_csv(here::here("data/dataset_1_tidy.csv"))

```


## Introduction

**What are the data i am looking at?**

- The data show the mean measured CO2 content of soil samples in different experiment 
sites. In all sites, there were two environments created. One with ambient CO2 
content in the atmosphere and one with elevated CO2 content in the atmosphere.
The experiments have been running for running for various amount of time.
The data also include the depth range of soil samples, the year and month of 
sampling, and source citations.

- **Source of the data:** 
Groenigen, Kees Jan van, Xuan Qi, Craig W. Osenberg, Yiqi Luo, and Bruce A. Hungate. 
“Faster Decomposition Under Increased Atmospheric CO2 Limits Soil Carbon Storage.
” Science 344, no. 6183 (May 2, 2014): 508–9. https://doi.org/10.1126/science.1249534. 

- The data was manually cleaned before my analysis.

**What do I expect my analysis to show, what is my hypothesis?**
**How should soil organic matter content change under elevated CO2?**

**My hypothesis:**

- Elevated CO2 content in the atmosphere will lead to an increase of soil organic 
matter over time. 

- Plants require CO2 for photosynthesis. Higher atmospheric CO2 levels enhance 
photosynthesis and therefore lead to an increase in productivity / more biomass. 
This biomass will over time fall off the plants and accumulate on the ground, 
leading to an increased soil organic matter content.

## Analysis


```{r}
# define early, mid and late phase since the start of the experiment
early_phase <- co2_impact_data |> filter(experiment_duration_years < 3)
mid_phase <- co2_impact_data |> filter(experiment_duration_years >= 3 & experiment_duration_years <= 6)
late_phase <- co2_impact_data |> filter(experiment_duration_years > 6)

# define a function that calculates the lrr
calculate_lrr <- function(ambient, increased){
  lrr <- (log(increased / ambient))
  return(lrr)
}

# calculate the lrr for every phase
early_phase$lrr <- calculate_lrr(early_phase$co2_ambient_mean, early_phase$co2_increased_mean)
mid_phase$lrr   <- calculate_lrr(mid_phase$co2_ambient_mean, mid_phase$co2_increased_mean)
late_phase$lrr  <- calculate_lrr(late_phase$co2_ambient_mean, late_phase$co2_increased_mean)

# calculate the mean lrr for each phase and save it in a data frame
lrr_summary <- data.frame(
  phase = c("early", "mid", "late"),
  mean_LRR = c(mean(early_phase$lrr, na.rm = TRUE),
               mean(mid_phase$lrr, na.rm = TRUE),
               mean(late_phase$lrr, na.rm = TRUE))
)

# present results
knitr::kable(lrr_summary, caption = "Mean LRR for each phase")

```


## Interpretation

**What do my final numbers mean?**

The final numbers are the mean LRR for each phase defined. 
This mean LRR shows how the soil organic content in the elevated CO2 environments 
changes in relation to the ambient CO2 environments.
Positive numbers indicate an increase in soil organic content in elevated CO2 
conditions
Negative numbers indicate a decrease in soil organic content in elevated CO2 
conditions.

The numbers do not support my initial hypothesis. 
I predicted that increased CO2 conditions in the atmosphere would lead to an 
increased soil organic content over time. But in the late phase, we observe a 
decrease in organic soil content in the elevated CO2 conditions relative to the 
ambient CO2 conditions. 
My observations after the analysis match the findings from the paper, where it 
was argued that increased CO2 levels also enhance decomposition of soil organic 
content. Thus, not leading to an increase of soil organic content over time.

